Section Two

### 2020-10-21 new2020/sections/sec02.Rmd
Some sec2 words

@@

## A transit road network

@@

The goal is to model vehicles as they travel along their respective routes, estimating their average speed along each road segment. Having traversed a road segment, the data is passed to the network to update the real-time traffic state. This can in turn be used, optionally with a forecasting function, to predict a future travel time of a bus along the road, which is used in arrival time prediction.

## Particle filters estimating vehicle state

@@

The particle filter allows us to estimate the state of a vehicle and, most importantly, it's average speed along road segments:
\[
   p(b_\ell | \vec y_{1:k}) \approx
   \sum_{i=1}^N w_{k}^{(i)} \DiracMeasure{b_{\ell}^{(i)}}{b_{\ell}}.
\]

## Estimating network state

@@

If a forecast function is available, we could also forecast future travel time based on current state and historical trends, and this would be incorporated.

>>>>

An important aspect missing from the setup, however, is information about *physical interactions* between separate routes (which may share common roads) which is critical to accurate prediction @citep:Jeong_2005,;. @citet:Elliott_2020; describe a simple process for combining route information from static GTFS data to obtain a *transit road network*, expressing each route as a sequence of physical roads. The benefit of this is that vehicles servicing different routes but using the same road(s) can share information about their speed. This greatly enhances the amount of data available for forecasting arrival times.

```{r road_nw,echo=FALSE}
# figure for the network
```

In the approach, stops are considered *nodes* in the network, while the roads between them are *edges*. Any buses travelling between the same two (or more) stops share the road between them, thus providing information about the speeds. This increases the rate of update and is particularly beneficial for predicting arrival times for low-frequency routes (see \cref{predicting-arrival-time}).

@@

<!-- ## A transit road network

@@

The goal is to model vehicles as they travel along their respective routes, estimating their average speed along each road segment. Having traversed a road segment, the data is passed to the network to update the real-time traffic state. This can in turn be used, optionally with a forecasting function, to predict a future travel time of a bus along the road, which is used in arrival time prediction. -->

## Recursive Bayes and particle filtering

In a real-time application, performing a full update of the posterior distribution of some parameters becomes hugely inefficient. It is possible to recursively update the posterior, however, by way of recursive Bayesian filtering (alternatively called sequential Monte Carlo). Given a random variable $\vec x = \vec x_{0:k} = \tvec{\vec x_0,\vec x_1,\ldots\,\vec x_k}$ at time $t_k$ following a Markov process with associated observation $\vec y = \vec y_{1:k} = \tvec{\vec y_1,\ldots,\vec y_k}$, the posterior distribution at time $t_k$ is expressed via the recursive formula
\[
p(\vec x_{0:k}\cond{} \vec y_{1:k}) \propto
   p(\vec x_{0:k-1}\cond{} \vec y_{1:k-1})
   p(\vec x_k\cond{}\vec x_{k-1})
   p(\vec y_k\cond{}\vec x_k)
\]
which involves the previous state, next state prediction, and the likelihood, respectively. We therefore need to compute $p(\vec x_k\cond{}\vec x_{k-1})$, which involves the *transition function* between consecutive states, and the likelihood $p(\vec y_k\cond{}\vec x_k)$. The specific details of each of these depends on the model, which is discussed in \cref{estimating-road-state-from-vehicle-observations}.

Estimation of recursive Bayesian models takes many forms, most notably the Kalman filter @citep:; which has been widely used in transit modelling @citep:Wall_1999,Dailey_2001,Cathey_2003,Jeong_2005;. However, as discussed by @citet:Hans_2015,Elliott_2020;, it makes several assumptions which do not work so well with transit vehicles, namely that the errors are normal with a single mode. The particle filter @citep:Gordon_1993; provides a numerical approach to solving recursive Bayesian models, much the same as Markov chain Monte Carlo provides estimates of standard posterior distributions. In a particle filter, the state $\vec x_k$ is represented by a sample of $N$\ particles $\{\vec x_k\vi\}_{i=1}^N$ with associated weights $\{w_k\vi\}_{i=1}^N$. We use the delta measure $\delta$ to express the approximated state
\[
   p(\vec x_{k-1} | \vec y_{0:k-1}) \approx \sum_{i=1}^N w_{k-1}^{(i)} \DiracMeasure{\vec x_{k-1}^{(i)}}{\vec x_{k-1}}.
\]

Particle filter estimates are updated in two steps. First the state prediction of $p(\vec x_k\cond{}\vec x_{k-1})$ is made using the transition function $f$, which implements the vehicle model (\cref{vehicle-model}):
\[
   p(\vec x_{k} | \vec x_{k-1}) \approx
      \sum_{i=1}^N w_{k-1}^{(i)}
         \DiracMeasure{\vec x_{k}^{(i)}}{\vec x_{k}},\quad
   \vec x_k\vi = f(\vec x_{k-1}\vi, \Delta_k, \sigma).
\]

Second is the likelihood update, which quantifies the probability of the observed vehicle position given each of the particle predictions. The particle weights are updated according to their likelihood,
\[
w_k\vi =
   \frac{w_{k-1}\vi p(\vec y_k\cond{}\vec x_k\vi)}{
      \sum_{j=1}^N w_{k-1}\vi[j] p(\vec y_k\cond{}\vec x_k\vi[j])
   }.
\]
For the likelihood function itself, we use an exponential distribution on the geographic distance between the particle and the observed location, as proposed by @citet:Elliott_2020;.

To ensure the *effective sample size* remains adequately large (greater than $\frac{1}{4}N)$, importance resampling of the particles, with replacemnet, is performed whenever the effective sample size $N_{\text{eff}} = \frac{1}{\sum_{i=1}^N (w_k\vi)^2}$ drops below the threshold. The particle filter was for this reason originally called the *bootstrap filter* @citep:Gordon_1993;.

## Estimating road state from vehicle observations

In a transit setup, we observe vehicle locations $\vec y_k$ consisting of the vehicle's latitude and longitude at time $t_k$. The goal is to estimate the vehicle's distance travelled $x$ and speed $\dot x$, from which we can calculate its average speed along individual road segments. The underlying, unobservable state of the vehicle is $\vec x_k = \tvec{x_k,\dots x_k}$.

Estimation of network state involves three steps: vehicle state estimation, estimation of average speed along roads, and updating the network state model.

### Vehicle model

Estimating vehicle state requires a vehicle model consisting of several key behaviours. @cite:Elliott_2020; give details, however of particular importance to arrival time prediction is the *dwell time* model for stop behaviour. When the bus arrives at a stop, it can either drive past, or stop and allow passengers to board and disembark.

<!--

@@

The particle filter allows us to estimate the state of a vehicle and, most importantly, it's average speed along road segments:
\[
   p(b_\ell | \vec y_{1:k}) \approx
   \sum_{i=1}^N w_{k}^{(i)} \DiracMeasure{b_{\ell}^{(i)}}{b_{\ell}}.
\] -->

<!--
## Estimating network state

@@

If a forecast function is available, we could also forecast future travel time based on current state and historical trends, and this would be incorporated. -->
